---
title: "EmissionsInContext"
author: "phoebe.woodworth-jefcoats@noaa.gov"
date: "2024-12-12"
output: html_document
---

This script documents and calculates the values in the Tech Memo's section that 
puts our emissions in a larger context.

```{r, echo = FALSE, message = FALSE}
# Set up the environment
library(tidyverse)
library(here)
```

```{r}
# Total and per capita emissions, from tech memo, in metric tons
PIR_emissions <- 6557.2 
PIR_percapita <- 22.5
```

```{r}
# State of Hawaiʻi emissions, in metric toms
HI_emissions <- 20.2e6 
HI_percapita <- 14.0
```


```{r}
# Values from World Bank data on countries':
# per capita emissions in t
# total emissions in kt

# Load data
WB_perCapita <- read_csv(here('Data', 'WorldBank', 'API_EN.ATM.CO2E.PC_DS2_en_csv_v2_637766.csv'), skip = 4,
                         show_col_type = FALSE)
WB_kt <- read_csv(here('Data', 'WorldBank', 'API_EN.ATM.CO2E.KT_DS2_en_csv_v2_637561.csv'), skip = 4,
                  show_col_types = FALSE)

# Condense down to just 2020
WB_perCapita_2020 <- WB_perCapita |> select(`Country Name`, `2020`)
WB_kt_2020 <- WB_kt |> select(`Country Name`, `2020`)

# Sort by 2020 value
WB_perCapita_2020_ranked <- arrange(WB_perCapita_2020, `2020`)
WB_kt_2020_ranked <- arrange(WB_kt_2020, `2020`)

# Lowest emitting countries
head(WB_kt_2020_ranked)

# Per capita
PIsmall <- WB_perCapita_2020_ranked |>
  filter(`Country Name` == 'Pacific island small states')
PIsmall
PIratio <- PIR_percapita/PIsmall$`2020`[1]

GT_PIR <- which(WB_perCapita_2020_ranked$`2020` > PIR_percapita)
WB_perCapita_2020_ranked[GT_PIR,]

US <- WB_perCapita_2020_ranked |>
  filter(`Country Name` == 'United States')
US
USratio <- PIR_percapita/US$`2020`[1]

HIratio <- PIR_percapita/HI_percapita
```
```{r}
# Volume of emissions
# derived using the ideal gas law: PV = nRT
P = 1 #atm
R = 0.0821 #L atm mol-1 K-1
T_std = 273.15 #K (0C)
T_room = 293.15 #K (20C)

CO2_mol = 12.011 + 2*15.9994 #g mol-1
n = 1e6 / CO2_mol #mol per ton CO2 (1e6 g per ton)

V_1ton_room_L <- (n * R * T_room) / P #L
V_1ton_room_m3 <- V_1ton_room_L * 0.001 #m3 (1 L = 0.001 m3)
V_PI_room_m3 <- V_1ton_room_m3 * PIR_emissions
Cube_edge <- V_PI_room_m3^(1/3)
```

```{r}
# Warming influence
# 1000 Gt CO2e = +0.45°C (IPCC)
# 1000 Gt = 1e12 t
PI_warming <- (PIR_emissions * 0.45) / 1e12
```

For Tech Memo:
Is 6,557.2 metric tons CO~2~, or 22.5 metric tons CO~2~e per employee, a lot?  Here are some ways to put these values in context:  

* In total, our operational emissions are comparable to that of the lowest-emitting 
Pacific Island nation, Tuvalu, and contribute roughly `r signif(PIR_emissions/HI_emissions*100, 1)`% 
of the total State of Hawaiʻi emissions (World Bank 2024a, State of Hawaiʻi 2024).  
* Per capita, our emissions are `r signif(PIratio, 2)` times 
those of Pacific Islands Small States, roughly `r signif(mean(c(USratio, HIratio)), 3)` 
times the average emissions from both the United States and the State of Hawaiʻi, 
and greater than all but a single nation globally (World Bank 2024b, State of Hawaiʻi 2024).  
* Our FY23 emissions would occupy `r signif(V_PI_room_m3/1e6, 3)` million cubic 
meters at room temperature, which is a cube that’s roughly `r signif(Cube_edge, 2)` meters on each side.  
* The warming influence of our FY23 operational emissions is `r signif(PI_warming*1e9, 3)` $\times$ 10^-^^9^ °C (IPCC 2023).
