---
title: "PIFSCtravelFeds"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2024-02-13"
output: html_document
---
# PIFSC official travel (federal staff) - emissions inventory
This document describes the methodology used to calculate the carbon footprint of PIFSC's official air
travel by federal staff.

### Data gathering
Travel data were obtained from the Director's Office via access to the FY23 Projected Travel Google Sheet (not linked, as it has controlled access).  Unnecessary data were redacted from the downloaded spreadsheet prior to working with the data.
The following data were retained:

- Traveler name (for determining travel origin; this is subsequently removed - see below)  
- Travel Description  
- Justification  
- Destination  
- Travel dates  

and saved as `FY23_PIFSC_Fed_Travel_Redacted_DIV.xlsx` where `DIV` is each of the six divisions that existed in FY23. 

```{r, include = FALSE}
# Set environment
library(tidyverse)
library(readxl)
library(here)
library(lubridate)
library(tidyxl)
```

```{r, include = FALSE}
# Handy info, uncomment if needed
TravelSheets <- excel_sheets(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted.xlsx'))
```

```{r, include = FALSE}
# Removing struck-through rows and reading in data, division by division
# Thanks to https://stackoverflow.com/questions/52107584/how-to-detect-strikethrough-style-from-xlsx-file-in-r

#-DO
formats_DO <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_DO.xlsx'))
cells_DO <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_DO.xlsx'))
strike_DO <- which(formats_DO$local$font$strike)
cells_DO_strike <- cells_DO[cells_DO$local_format_id %in% strike_DO, 2]
cells_DO_strike <- parse_number(cells_DO_strike$address)
DO_rows_strike <- unique(cells_DO_strike)
AllTravel_DO <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_DO.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))
# Remove struck-through rows, subtracting 1 to account for the header
Travel_DO <- AllTravel_DO[-(DO_rows_strike - 1),]

#-ESD
formats_ESD <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_ESD.xlsx'))
cells_ESD <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_ESD.xlsx'))
strike_ESD <- which(formats_ESD$local$font$strike)
cells_ESD_strike <- cells_ESD[cells_ESD$local_format_id %in% strike_ESD, 2]
cells_ESD_strike <- parse_number(cells_ESD_strike$address)
ESD_rows_strike <- unique(cells_ESD_strike)
AllTravel_ESD <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_ESD.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))
# Remove struck-through rows, subtracting 1 to account for the header
Travel_ESD <- AllTravel_ESD[-(ESD_rows_strike - 1),]

#-FRMD
formats_FRMD <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_FRMD.xlsx'))
cells_FRMD <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_FRMD.xlsx'))
strike_FRMD <- which(formats_FRMD$local$font$strike)
cells_FRMD_strike <- cells_FRMD[cells_FRMD$local_format_id %in% strike_FRMD, 2]
cells_FRMD_strike <- parse_number(cells_FRMD_strike$address)
FRMD_rows_strike <- unique(cells_FRMD_strike)
AllTravel_FRMD <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_FRMD.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))
# Remove struck-through rows, subtracting 1 to account for the header
Travel_FRMD <- AllTravel_FRMD[-(FRMD_rows_strike - 1),]

#-OMI
formats_OMI <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_OMI.xlsx'))
cells_OMI <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_OMI.xlsx'))
strike_OMI <- which(formats_OMI$local$font$strike)
cells_OMI_strike <- cells_OMI[cells_OMI$local_format_id %in% strike_OMI, 2]
cells_OMI_strike <- parse_number(cells_OMI_strike$address)
OMI_rows_strike <- unique(cells_OMI_strike)
AllTravel_OMI <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_OMI.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))
# Remove struck-through rows, subtracting 1 to account for the header
Travel_OMI <- AllTravel_OMI[-(OMI_rows_strike - 1),]

#-PSD
formats_PSD <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_PSD.xlsx'))
cells_PSD <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_PSD.xlsx'))
strike_PSD <- which(formats_PSD$local$font$strike)
cells_PSD_strike <- cells_PSD[cells_PSD$local_format_id %in% strike_PSD, 2]
cells_PSD_strike <- parse_number(cells_PSD_strike$address)
PSD_rows_strike <- unique(cells_PSD_strike)
AllTravel_PSD <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_PSD.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))
# Remove struck-through rows, subtracting 1 to account for the header
Travel_PSD <- AllTravel_PSD[-(PSD_rows_strike - 1),]

#-SOD
formats_SOD <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_SOD.xlsx'))
cells_SOD <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_SOD.xlsx'))
strike_SOD <- which(formats_SOD$local$font$strike)
cells_SOD_strike <- cells_SOD[cells_SOD$local_format_id %in% strike_SOD, 2]
cells_SOD_strike <- parse_number(cells_SOD_strike$address)
SOD_rows_strike <- unique(cells_SOD_strike)
AllTravel_SOD <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_SOD.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))
# Remove struck-through rows, subtracting 1 to account for the header
Travel_SOD <- AllTravel_SOD[-(SOD_rows_strike - 1),]

# Find Total Travel events
TotalEvents <- dim(AllTravel_DO)[1] + 
  dim(AllTravel_ESD)[1] + 
  dim(AllTravel_FRMD)[1] + 
  dim(AllTravel_OMI)[1] + 
  dim(AllTravel_PSD)[1] + 
  dim(AllTravel_SOD)[1]

# Find total of struck-through entries
TotalStruckThrough <- length(DO_rows_strike) + 
  length(ESD_rows_strike) + 
  length(FRMD_rows_strike) + 
  length(OMI_rows_strike) + 
  length(PSD_rows_strike) + 
  length(SOD_rows_strike)

# Concatenate
AllTravel <- bind_rows(Travel_DO, Travel_ESD, Travel_FRMD, Travel_OMI, Travel_PSD, AllTravel_SOD)
```

```{r, include = FALSE}
# Remove rows with missing travel start and/or end dates
NoTravelDate <- which(is.na(AllTravel$`TRAVEL\r\nSTART`) | is.na(AllTravel$`TRAVEL\r\nEND`))
AllTravel <- AllTravel[-NoTravelDate,]
```

```{r, include = FALSE}
# Remove non-specific destinations
DestNA <- which(is.na(AllTravel$DESTINATION))
AllTravel <- AllTravel[-DestNA,]
DestTBD_idx <- str_detect(AllTravel$DESTINATION, 'TBD')
DestTBD <- which(DestTBD_idx == TRUE)
AllTravel <- AllTravel[-DestTBD,]
```

```{r, include = FALSE}
# Remove local travel
DestLocal_idx <- str_detect(AllTravel$DESTINATION, fixed('local', ignore_case = TRUE))
DestLocal <- which(DestLocal_idx == TRUE)
AllTravel <- AllTravel[-DestLocal,]

NotesLocal_idx <- str_detect(AllTravel$NOTES, fixed('local', ignore_case = TRUE))
NotesLocal <- which(NotesLocal_idx == TRUE)
# AllTravel <- AllTravel[-NotesLocal,] # Empty

JustLocal_idx <- str_detect(AllTravel$JUSTIFICATION, fixed('local', ignore_case = TRUE))
JustLocal <- which(JustLocal_idx == TRUE)
AllTravel <- AllTravel[-JustLocal,]

TotalLocal <- length(DestLocal) + length(NotesLocal) + length(JustLocal)
```

```{r, include = FALSE}
# Save the initial file from which we're going to manually remove the things discussed below.
AllTravel <- as_tibble(AllTravel)
write_csv(AllTravel, file = here('Data', 'AirTravel', 'FY23_PIFSC_Fed_Travel_Redacted_Filtered.csv'))
```

```{r, include = FALSE}
# Count and remove the cruises, which were manually identified and struckthrough
formats_cruise <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_RemoveCruises.xlsx'))
cells_cruise <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_RemoveCruises.xlsx'))
strike_cruise <- which(formats_cruise$local$font$strike)
cells_cruise_strike <- cells_cruise[cells_cruise$local_format_id %in% strike_cruise, 2]
cells_cruise_strike <- parse_number(cells_cruise_strike$address)
cruise_rows_strike <- unique(cells_cruise_strike)
AllTravel_cruise <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_RemoveCruises.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text'))

# Remove struck-through rows, subtracting 1 to account for the header
Travel_noCruise <- AllTravel_cruise[-(cruise_rows_strike - 1),]

# Save, for removing local travel
Travel_noCruise <- as_tibble(Travel_noCruise)
write_csv(Travel_noCruise, file = here('Data', 'AirTravel', 'FY23_PIFSC_Fed_Travel_Redacted_NoCruise.csv'))
```

```{r, include = FALSE}
# Count and remove travel to/from Honolulu for:
# - Staff whom the NOAA directory says are based at IRC
# - Individuals who don't appear in the NOAA directory
# Commented out because the code below was used to ID rows 
# and a subsequent column was added
# Plus, some of the ID'd rows have been removed by the process on lines 234 & 235
# AllTravel_wIRCbased <- read_excel(here('Data', 'AirTravel', 
#                              'FY23_PIFSC_Fed_Travel_Redacted_NoHNLforIRCbased.xlsx'),
#                            col_types = c('text', 'text', 'text', 'text', 
#                                       'date', 'date', 'text'))
# 
# hon_idx <- str_detect(AllTravel_wIRCbased$DESTINATION, fixed('hon', ignore_case = TRUE))
# hon <- which(hon_idx == TRUE)
# hnl_idx <- str_detect(AllTravel_wIRCbased$DESTINATION, fixed('hnl', ignore_case = TRUE))
# hnl <- which(hnl_idx == TRUE)

# Remove records as explained above
# Manually add 'ORIGIN' column as needed

# Now load the manually edited file to count removed records
# Note that we added an 'ORIGIN' column
formats_hon <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_NoHNLforIRCbased.xlsx'))
cells_hon <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_NoHNLforIRCbased.xlsx'))
strike_hon <- which(formats_hon$local$font$strike)
cells_hon_strike <- cells_hon[cells_hon$local_format_id %in% strike_hon, 2]
cells_hon_strike <- parse_number(cells_hon_strike$address)
hon_rows_strike <- unique(cells_hon_strike)
AllTravel_hon <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_NoHNLforIRCbased.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text', 'text'))

# Remove struck-through rows, subtracting 1 to account for the header
Travel_noHon <- AllTravel_hon[-(hon_rows_strike - 1),]

# Save this, and then we'll (finally) manually remove the duplicates (and be done with cleaning the data)
Travel_noHon <- as_tibble(Travel_noHon)
write_csv(Travel_noHon, file = here('Data', 'AirTravel', 'FY23_PIFSC_Fed_Travel_Redacted_NoHon.csv'))
```

```{r, include = FALSE}
# Using 'Travel_noHon' to:
# - Identify unique travelers
# - Identify duplicate travel
# - Strike through duplicates in 'FY23_PIFSC_Fed_Travel_Redacted_NoDups.xlsx'

travelers <- unique(Travel_noHon$`TRAVELER NAME                      (LAST, FIRST)`)

# This is tedious....
# Changing the traveler number manually
# Commented out to suppress output
tv <- which(Travel_noHon$`TRAVELER NAME                      (LAST, FIRST)` == travelers[91])
# Travel_noHon[tv,]
```

```{r, include = FALSE}
# Nearly there....  
# Load the file with duplicates struck through
# Remove them
# Add in Honolulu as ORIGIN where it's missing
# Replace travelers with unique IDs
# Save this file to work with
# Whew.  Time for a snack.
formats_NoDups <- xlsx_formats(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_NoDups.xlsx'))
cells_NoDups <- xlsx_cells(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_NoDups.xlsx'))
strike_NoDups <- which(formats_NoDups$local$font$strike)
cells_NoDups_strike <- cells_NoDups[cells_NoDups$local_format_id %in% strike_NoDups, 2]
cells_NoDups_strike <- parse_number(cells_NoDups_strike$address)
NoDups_rows_strike <- unique(cells_NoDups_strike)
AllTravel_NoDups <- read_excel(here('Data', 'AirTravel', 
                             'FY23_PIFSC_Fed_Travel_Redacted_NoDups.xlsx'),
                           col_types = c('text', 'text', 'text', 'text', 
                                      'date', 'date', 'text', 'text'))

# Remove struck-through rows, subtracting 1 to account for the header
AirTravel <- AllTravel_NoDups[-(NoDups_rows_strike - 1),]

# Fill in 'Honolulu, HI'  where origin isn't specified
# Noting that this column is read in as text
HNL <- str_detect(AirTravel$ORIGIN, fixed('NA'))
AirTravel$ORIGIN[which(HNL == TRUE)] <- 'Honolulu, HI'

# Finally, save a version where names are replaced by unique numeric identifiers
n_travelers <- unique(AirTravel$`TRAVELER NAME                      (LAST, FIRST)`)
uq_id <- sample.int(length(n_travelers), length(n_travelers), replace = FALSE)

TravelerID <- matrix(NA, nrow = nrow(AirTravel))
for (r in seq(1, length(TravelerID), 1)) {
  idx <- which(n_travelers == AirTravel$`TRAVELER NAME                      (LAST, FIRST)`[r])
  TravelerID[r,1] <- uq_id[idx]
}

# Add this to the matrix, remove the Traveler Name column, and save
AirTravel_Unique <- bind_cols(TravelerID,
                              AirTravel$`CONFERENCE/MEETING NAME, TRAVEL DESCRIPTION`,
                              AirTravel$JUSTIFICATION,
                              AirTravel$DESTINATION,
                              AirTravel$`TRAVEL\r\r\r\r\nSTART`,
                              AirTravel$`TRAVEL\r\r\r\r\nEND`,
                              AirTravel$NOTES,
                              AirTravel$ORIGIN)
AirTravel_Unique <- rename(AirTravel_Unique, TravelerID = ...1)
AirTravel_Unique <- rename(AirTravel_Unique, Description = ...2)
AirTravel_Unique <- rename(AirTravel_Unique, Justification = ...3)
AirTravel_Unique <- rename(AirTravel_Unique, Destination = ...4)
AirTravel_Unique <- rename(AirTravel_Unique, TravelStart = ...5)
AirTravel_Unique <- rename(AirTravel_Unique, TravelEnd = ...6)
AirTravel_Unique <- rename(AirTravel_Unique, Notes = ...7)
AirTravel_Unique <- rename(AirTravel_Unique, Origin = ...8)

# Save this, and then we'll (finally) manually remove the duplicates (and be done with cleaning the data)
write_csv(AirTravel_Unique, file = here('Data', 'AirTravel', 'FY23_PIFSC_Fed_AirTravel_Redacted.csv'))
```

### Data wrangling and assumptions
To get a preliminary estimate of FY23 emissions from PIFSC air travel in FY23, considerable manual data wrangling was needed.  From `FY23_PIFSC_Fed_Travel_Redacted.xlsx`, rows preceding the header row were removed.  Rows identifying and separating quarters were also removed, as were summary and instructional rows.  When attempting to load the `FY23_PIFSC_Fed_Travel_Redacted_DIV.xlsx` files into R numerous records triggered date-related warning messages.  These were edited by hand to alter the date format while preserving the date information. Most commonly, this required editing 2-digit years (e.g., 23) to be 4-digit years (e.g., 2023), adding in missing years, adding leading zeros to single-digit months, changing the Excel cell classification from 'Text' to 'Short Date,' and rewriting text dates as numeric dates (e.g., April 30 to 04/30/2023).  For struck-through rows, problematic dates were manually deleted.  Vague dates (e.g., March 2023) were manually deleted and entries counted as below.

Traveler names were manually edited to be in the format of Last, First where they were written otherwise.  First names were edited to be written in their entirety where multiple versions were presented.  This was done in order to accurately replace travelers' identities with unique identifiers.  In some cases, a single traveler's name was entered different ways for different entries (e.g, Last, First as well as First Last; Doe, Jon as well as Doe, Jonathan).  

Data were pared down to only air travel 
(`FY23_PIFSC_Fed_AirTravel_Redacted.xlsx`) by:  

* Removing any records that included the words 'local travel' (_n_ = `r TotalLocal`)  
* Manually removing ship-based travel (which is captured elsewhere in the inventory).  It is possible that by
removing ship-based travel that didn't clearly include a leg of air travel, instances in which travelers are flown to or from the vessel may be missed.  Also, there were two records where it clearly stated that staff travel between Honolulu and Pago Pago was by by ship one way and by airplane one way.  Because subsequent methodology treats trips as round trips, one record was kept and one record was manually omitted.  (_n_ = `r length(cruise_rows_strike)`)    
* Manually removing records with a destination of 'Honolulu' or 'Main Hawaiian Islands' for IRC-based staff. (_n_ = `r length(hon_rows_strike)`)

Names were replaced by unique identifiers.  

Additionally:  

* Duplicate travel events were manually removed (_n_ = `r length(NoDups_rows_strike)`)    
* Travel events with 'tbd' destinations were removed (_n_ = `r length(DestTBD)`)    
* Travel events with no destination were removed (_n_ = `r length(DestNA)`)    
* Travel events without specific travel start and travel end dates were removed (_n_ = `r length(NoTravelDate)`)    

The following assumptions were made about the travel data:    

* Honolulu was assumed to be the travel origin city unless it was listed as the destination and not
flagged as 'local travel' and the traveler was not based at IRC.  In these instances, city listed 
for the traveler's NOAA Staff Directory was used. Travel to/from Honolulu for remote staff with Honolulu listed as their address in the NOAA Staff Directory was omitted due to lack of information on travel origin.  
* Struck-through entries were assumed not to have taken place (_n_ = `r  TotalStruckThrough`).    

All other travel entered was assumed to have be carried out and treated as described above.

After applying the above filters and assumptions, `r dim(AirTravel)[1]` out of an initial `r TotalEvents` individual trips remained.  Note that this would translate to `r (dim(AirTravel)[1]) * 2` unique travel events because each trip is treated as round-trip.

These were further condensed by assuming:  

* travelers traveling on the same dates to and from the same origin and destination were traveling on the same flights.  

``` {r, include = FALSE}
# First, let's clean up and load only the file we need to move ahead.
rm(list = ls())

AirTravel <- read_csv(here('Data', 'AirTravel', 'FY23_PIFSC_Fed_AirTravel_Redacted.csv'))
```

```{r, include = FALSE}
# Identify unique Origin and Destination pairs
indvOrigDest <- unique(AirTravel[,c('Destination', 'Origin')])

# For each destination:
# Find each unique origin
# Fine unique travel start dates
# Find unique travel end dates
# Append to matrix that will have all unique travel events

d_hold <- filter(AirTravel, 
                (Destination == indvOrigDest$Destination[1] & 
                 Origin == indvOrigDest$Origin[1]))

n_begin <- unique(d_hold$`TravelStart`)
n_end <- unique(d_hold$`TravelEnd`)
 
OutboundBeginLocation <- rep(d_hold$Origin[1], length(n_begin))
OutboundEndLocation <- rep(d_hold$Destination[1], length(n_begin))
InboundBeginLocation <- rep(d_hold$Destination[1], length(n_end))
InboundEndLocation <- rep(d_hold$Origin[1], length(n_end))
 
Outbound <- bind_cols(OutboundBeginLocation, OutboundEndLocation)
Inbound <- bind_cols(InboundBeginLocation, InboundEndLocation)
 
FlightLegs <- bind_rows(Outbound, Inbound)

for (d in seq(2, nrow(indvOrigDest), 1)) {
 d_hold <- filter(AirTravel, 
                  (Destination == indvOrigDest$Destination[d] & 
                   Origin == indvOrigDest$Origin[d]))
 
 n_begin <- unique(d_hold$`TravelStart`)
 n_end <- unique(d_hold$`TravelEnd`)
 
 OutboundBeginLocation <- rep(d_hold$Origin[1], length(n_begin))
 OutboundEndLocation <- rep(d_hold$Destination[1], length(n_begin))
 InboundBeginLocation <- rep(d_hold$Destination[1], length(n_end))
 InboundEndLocation <- rep(d_hold$Origin[1], length(n_end))
 
 Outbound <- bind_cols(OutboundBeginLocation, OutboundEndLocation)
 Inbound <- bind_cols(InboundBeginLocation, InboundEndLocation)
 
 FlightLegs <- bind_rows(FlightLegs, Outbound, Inbound)
}
 
FlightLegs <- rename(FlightLegs, BeginLocation = ...1)
FlightLegs <- rename(FlightLegs, EndLocation = ...2)

write_csv(FlightLegs, here('Data', 'AirTravel', 'FY23_PIFSC_FlightLegs.csv'))
 
# Identify all unique cities
# Commented out to avoid overwriting file with Lat & Lon added
# AllCities <- bind_rows(FlightLegs[,1], FlightLegs[,2])
# AllFlightCities <- unique(AllCities[,1])
# AllFlightCities <- rename(AllFlightCities, City = BeginLocation)
# # write_csv(AllFlightCities, here("Data", 'UniqueFlightLocations.csv'))
 
# Load city locations
FlightCities <- read_csv(here('Data', 'AirTravel', "UniqueFlightLocations.csv"))

# For each flight leg:
# Add BeginLocation lat/lon -> Blat/Blon
# Add EndLocation lat/lon -> Elat/Elon
Blat <- matrix(NA, nrow = nrow(FlightLegs), ncol = 1)
Blon <- matrix(NA, nrow = nrow(FlightLegs), ncol = 1)
Elat <- matrix(NA, nrow = nrow(FlightLegs), ncol = 1)
Elon <- matrix(NA, nrow = nrow(FlightLegs), ncol = 1)
for (c in seq(1, nrow(FlightLegs), 1)) {
  Bidx <- which(FlightCities$City == FlightLegs$BeginLocation[c])
  Eidx <- which(FlightCities$City == FlightLegs$EndLocation[c])

  Blat[c,1] <- FlightCities$Lat[Bidx]
  Blon[c,1] <- FlightCities$Lon[Bidx]
  Elat[c,1] <- FlightCities$Lat[Eidx]
  Elon[c,1] <- FlightCities$Lon[Eidx]
}

FlightLegs <- bind_cols(FlightLegs, Blat, Blon, Elat, Elon)
FlightLegs <- rename(FlightLegs, BeginLat = ...3)
FlightLegs <- rename(FlightLegs, BeginLon = ...4)
FlightLegs <- rename(FlightLegs, EndLat = ...5)
FlightLegs <- rename(FlightLegs, EndLon = ...6)

# This approach (https://edwilliams.org/avform147.htm, linked from NOAA's NHC) uses
# N and W as positive and S and E as negative, so (sigh):
Blon_conv <- Blon
Elon_conv <- Elon
for (r in seq(1, length(Blon), 1)) {
  if (Blon_conv[r,1] < 180) {
    Blon_conv[r,1] = Blon[r,1] * -1
  } else if (Blon_conv[r,1] > 180) {
    Blon_conv[r,1] = 360 - Blon[r,1]
  }

  if (Elon_conv[r,1] < 180) {
    Elon_conv[r,1] = Elon[r,1] * -1
  } else if (Elon_conv[r,1] > 180) {
    Elon_conv[r,1] = 360 - Elon[r,1]
  }
}

# It also uses radians rather than degrees
# And nm rather than statue miles
deg2rad <- function(deg) {
  return((pi/180) * deg)
}

FlightDistance <- matrix(NA, nrow = length(Blat))
for (r in seq(1, length(FlightDistance), 1)) {
  FlightDistance[r] <- acos(sin(deg2rad(Blat[r])) *
                            sin(deg2rad(Elat[r])) +
                            cos(deg2rad(Blat[r])) *
                            cos(deg2rad(Elat[r])) *
                            cos(deg2rad(Blon_conv[r]) - deg2rad(Elon_conv[r]))) *
                        ((180*60)/pi) * 1.15078
}

FlightLegs <- bind_cols(FlightLegs, FlightDistance)
FlightLegs <- rename(FlightLegs, FlightDistance = ...7)

write_csv(FlightLegs, here('Data', 'AirTravel', "FY23_PIFSC_UniqueFlightLegs.csv"))
```

The final list of air travel trips included `r dim(FlightLegs)[1]` unique travel events (outbound and inbound trips are listed separately).  These travel events were saved to the file `FY23_PIFSC_UniqueFlightLegs.csv`.

The following location assumptions were made:  

* Both 'Guam and Saipan' and 'Guam, Saipan' were assumed to be 'Guam'  
* 'Big Island' was assumed to be the island of Hawaiʻi  
* 'Tutuila; Ofu; Ta'u' was assumed to be 'Tutuila'  
* 'Perth, Western Australia/Hobart, Tasmania' was assumed to be 'Perth, Western Australia'  
* 'Kei and Buru islands, Bali Straits' was assumed to be 'Kei islands, Indonesia'  
* 'Rose Atoll, American Samoa' was assumed to be 'Rose Atoll'  
* 'Honolulu, Kona' was assumed to be 'Honolulu'  
* various misspellings were assumed to be the corrected result returned by Google  

A list of all unique locations (origins and destinations) was generated (`UniqueFlightLocations.csv`; note that these are unique character strings, which include multiple representations of the same location).  To this list, latitudes and longitudes were added manually by searching for city (or island, state, or country, if that's all that was provided) locations on Google.  West longitudes were converted to 360-degree longitude by subtracting the value from 360.  For Honolulu, the Honolulu airport location was used.  

### Analysis
Distances were measured using the same approach as used by the NOAA National Hurricane Center's [Latitude/Longitude Distance Calculator](https://www.nhc.noaa.gov/gccalc.shtml): Ed Williams's [Aviation Formulary V1.47](https://edwilliams.org/avform147.htm).

Rather than copy and paste all `r dim(FlightLegs)[1]` unique flight legs in the EPA's calculator tool, and manually selecting the flight length category for each (short, medium, or long haul), the following approach from the calculator was used in R, using the following emissions factors:  

* Air Short Haul (< 300 miles):  
     * CO~2~ Factor = 0.207 kg per passenger-mile  
     * CH~4~ Factor = 0.0064 g per passenger-mile  
     * N~2~O Factor = 0.0066 g per passenger-mile  
* Air Medium Haul (>= 300 miles, < 2300 miles):   
     * CO~2~ Factor = 0.129 kg per passenger-mile  
     * CH~4~ Factor = 0.0006 g per passenger-mile  
     * N~2~O Factor = 0.0041 g per passenger-mile  
* Air Long Haul (>= 2300 miles):   
     * CO~2~ Factor = 0.163 kg per passenger-mile  
     * CH~4~ Factor = 0.0006 g per passenger-mile  
     * N~2~O Factor = 0.0052 g per passenger-mile  

Emissions for each gas were determined by multiplying the appropriate factor by the distance traveled.  Emissions were summed for each gas, multiplied by the appropriate Global Warming Potential (GWP) factor:  

* CO~2~ GWP = 1  
* CH~4~ GWP = 25  
* N~2~O GWP = 298  

and converted to metric tons.

```{r, include = FALSE}
# Clean up again and use only what we need
rm(list = ls())

FlightLegs <- read_csv(here('Data', 'AirTravel', "FY23_PIFSC_UniqueFlightLegs.csv"))

# Calculate emissions
CO2emissions <- matrix(NA, nrow = dim(FlightLegs)[1])
CH4emissions <- matrix(NA, nrow = dim(FlightLegs)[1])
N2Oemissions <- matrix(NA, nrow = dim(FlightLegs)[1])

ShortHaul <- which(FlightLegs$FlightDistance < 300)
MedHaul <- which(FlightLegs$FlightDistance >= 300 & FlightLegs$FlightDistance < 2300)
LongHaul <- which(FlightLegs$FlightDistance >= 2300)

CO2emissions[ShortHaul] = FlightLegs$FlightDistance[ShortHaul] * 0.207
CO2emissions[MedHaul] = FlightLegs$FlightDistance[MedHaul] * 0.129
CO2emissions[LongHaul] = FlightLegs$FlightDistance[LongHaul] * 0.163

CH4emissions[ShortHaul] = FlightLegs$FlightDistance[ShortHaul] * 0.0064
CH4emissions[MedHaul] = FlightLegs$FlightDistance[MedHaul] * 0.0006
CH4emissions[LongHaul] = FlightLegs$FlightDistance[LongHaul] * 0.0006

N2Oemissions[ShortHaul] = FlightLegs$FlightDistance[ShortHaul] * 0.0066
N2Oemissions[MedHaul] = FlightLegs$FlightDistance[MedHaul] * 0.0041
N2Oemissions[LongHaul] = FlightLegs$FlightDistance[LongHaul] * 0.0052

FlightEmissions <- (sum(CO2emissions) +
                    sum(CH4emissions) * 25 / 1000 +
                    sum(N2Oemissions) * 298 / 1000) / 1000
```

## Preliminary results
The preliminary estimate is that in FY23, PIFSC's official air travel emitted the equivalent of `r signif(FlightEmissions,4)` metric tons of CO~2~.  

## Known limitations
This approach assumes that all trips are a single point-to-point flight.  It is virtually certain that this underestimates the total emissions by underestimating the total number of flights and the total miles flown.

In at least one instance, the dates in the travel spreadsheet failed to take into account personal leave appended to official travel.  This resulted in the assumption that multiple travelers were on a single flight, when in fact they travel separately.  If this is widespread across the travel spreadsheet, emissions are underestimated.

All trips were assumed to be round trips.  In some cases, one direction was made via ship rather than air.  This results in an overestimate of flight-related emissions.  Based on comments in the spreadsheet, this likely to affects < 10% of all trips.  Similarly, there were several trips where travel dates were adjacent to another trip made by the same traveler, indicating that it's possible a multi-leg trip was taken rather than two round trips.  Likewise, this methodology likely omits additional legs for trips with multiple destinations.

```{r, include = FALSE}
# # Error check - commented out to preserve results
# L10 <- sample.int(dim(FlightLegs)[1], 10, replace = FALSE)
# # Saving this so it can be replicated
# DistanceCheck <- bind_cols(FlightLegs$BeginLat[L10],
#                            FlightLegs$BeginLon[L10],
#                            FlightLegs$EndLat[L10],
#                            FlightLegs$EndLon[L10],
#                            FlightLegs$FlightDistance[L10])
# DistanceCheck <- rename(DistanceCheck, BeginLat = ...1)
# DistanceCheck <- rename(DistanceCheck, BeginLon = ...2)
# DistanceCheck <- rename(DistanceCheck, EndLat = ...3)
# DistanceCheck <- rename(DistanceCheck, EndLon = ...4)
# DistanceCheck <- rename(DistanceCheck, FlightDistance_Methods = ...5)
# 
# ManualResults <- c(3853, 6771, 2593, 2534, 2680, 5846, 3849, 2561, 5616, 5081)
# ManualResults <- as.matrix(ManualResults)
# 
# DistanceCheck <- bind_cols(DistanceCheck, ManualResults)
# DistanceCheck <- rename(DistanceCheck, FlightDistance_Check = ...6)
# write_csv(DistanceCheck, here('Data', 'AirTravel', "PIFSC_FlightEmissions_DistanceCheck.csv"))

# # Error check - commented out to preserve results
# EC1 <- sample.int(dim(FlightLegs)[1], 32, replace = FALSE)
# EC2 <- sample.int(dim(FlightLegs)[1], 32, replace = FALSE)
# EC3 <- sample.int(dim(FlightLegs)[1], 32, replace = FALSE)
# # Saving these so that this can be replicated
# EC1 <- as.matrix(EC1)
# EC2 <- as.matrix(EC2)
# EC3 <- as.matrix(EC3)
# ErrorChecks <- bind_cols(EC1, EC2, EC3)
# write_csv(ErrorChecks, here('Data', 'AirTravel', "PIFSC_FlightEmissions_ErrorChecks.csv"))
# 
# Test1 <- FlightLegs$FlightDistance[EC1]
# Test2 <- FlightLegs$FlightDistance[EC2]
# Test3 <- FlightLegs$FlightDistance[EC3]
# 
# 
# Sub1 <- (sum(CO2emissions[EC1]) +
#          sum(CH4emissions[EC1]) * 25 / 1000 +
#          sum(N2Oemissions[EC1]) * 298 / 1000) / 1000
# 
# Sub2 <- (sum(CO2emissions[EC2]) +
#          sum(CH4emissions[EC2]) * 25 / 1000 +
#          sum(N2Oemissions[EC2]) * 298 / 1000) / 1000
# 
# Sub3 <- (sum(CO2emissions[EC3]) +
#          sum(CH4emissions[EC3]) * 25 / 1000 +
#          sum(N2Oemissions[EC3]) * 298 / 1000) / 1000
```

## Error checking
Ten random  travel legs were selected and their distance was computed using the NHC's lat/lon distance calculator for comparison to our methods to ensure the calculation was correctly coded.  All ten distances compared were within one mile of the NHC's calculator which rounds to the nearest mile. 


The default version of the EPA's calculator tool has 32 rows for air travel.  To spot check
the calculations used here, 32 random flight legs were selected.  The emissions for this subset was calculated using the above methodology and using the calculator tool.  This was repeated three times.  For the first iteration, the calculator tool estimated 18.7 metric tons CO~2~e and this methodology estimated 18.7 tons CO~2~e.  For the second iteration, the calculator tool estimated 16.2 tons CO~2~e and this methodology estimated 16.2 tons CO~2~e.  For the third iteration, the calculator tool estimated 20.3 tons CO~2~e and this methodology estimated 20.3 tons CO~2~e.

```{r, include = FALSE}
# Scaling up the error checkestimates to the full number of flight legs
pct_samp <- 32/dim(FlightLegs)[1]*100

sup1 <- 18.7 / (32/dim(FlightLegs)[1])
sup2 <- 16.2 / (32/dim(FlightLegs)[1])
sup3 <- 20.3 / (32/dim(FlightLegs)[1])

avg_scale <- (sup1 + sup2 + sup3)/3
```

Additionally, each of the three error checks sampled `r signif(pct_samp, 3)`% of the flights.  Scaling these estimates up results in `r signif(sup1, 4)` tons CO~2~e, `r signif(sup2, 4)` tons CO~2~e, and `r signif(sup3, 4)` tons CO~2~e respectively, which averages to `r signif(avg_scale, 4)` tons CO~2~e and is quite close to our calculated total of `r signif(FlightEmissions,4)` tons CO~2~e.

## Final estimate
Accepting the known limitations described above, the final estimate is that in FY23 PIFSC's official air travel emitted the equivalent of `r signif(FlightEmissions,4)` metric tons of CO~2~.