---
title: "PIFSCrefrigerants"
output: html_document
date: "2024-08-03"
---

# PIFSC refrigerators and freezers
PIFSC has a number of refrigerators, freezers, and other appliances that use refrigerants.  

## Methods
This section describes the methods used to estimate the emissions from PIFSC refrigerant usage, which follow those in the US EPA’s calculator tool.

### Data Gathering, Wrangling, and Assumptions
A list of cold storage units and other appliances that use refrigeration was provided by Kerry Reardon (Kerry.Reardon@noaa.gov).  This information is accessible via Google Drive and a copy has been added to our Data folder as both a Google Sheet and Excel spreadsheet (to make analysis easier).

```{r, include = FALSE}
# Set environment
library(tidyverse)
library(readxl)
library(here)
library(lubridate)
library(tidyxl)
```

```{r, include = FALSE}
# Load data
Refrigerators <- read_xlsx(here("Data", "Copy of List of cold storage units and large appliances 2024.07.30.xlsx"), 
                           sheet = "Cold Storage")
Reefers_with <- filter(Refrigerators, !is.na(Refrigerant))

MiscApps <- read_xlsx(here("Data", "Copy of List of cold storage units and large appliances 2024.07.30.xlsx"), 
                           sheet = "MISC large appliances")
MiscApps_with <- MiscApps |> filter(if_all(everything(), ~!is.na(Refrigerant)))
```

```{r, echo = FALSE}
# Identify all refrigerants used and assign their global warming potential (GWP)
Refs_reefers_1 <- unique(Reefers_with$Refrigerant)
Refs_misc_1 <- unique(MiscApps_with$Refrigerant)
Refs_used_1 <- unique(c(Refs_reefers_1, Refs_misc_1))

# Some units use multiple refrigerants.
# We're going to break these into duplicate entries with each refrigerant listed separately
Mult_1 <- which(Reefers_with$Refrigerant == "R404A, R170, R508B")
Mult_1_R404A <- Reefers_with[Mult_1,]
Mult_1_R404A$Refrigerant <- "R404A"
Mult_1_R170 <- Reefers_with[Mult_1,]
Mult_1_R170$Refrigerant <- "R170"
Mult_1_R508B <- Reefers_with[Mult_1,]
Mult_1_R508B$Refrigerant <- "R508B"
Reefers_single_round1 <- Reefers_with[-Mult_1,]
Reefers_single_round1 <- bind_rows(Reefers_single_round1, Mult_1_R404A, Mult_1_R170, Mult_1_R508B)

Mult_2 <- which(Reefers_single_round1$Refrigerant == "R134A, R404A, R170, R508B")
Mult_2_R134A <- Reefers_single_round1[Mult_2,]
Mult_2_R134A$Refrigerant <- "R134A"
Mult_2_R404A <- Reefers_single_round1[Mult_2,]
Mult_2_R404A$Refrigerant <- "R404A"
Mult_2_R170 <- Reefers_single_round1[Mult_2,]
Mult_2_R170$Refrigerant <- "R170"
Mult_2_R508B <- Reefers_single_round1[Mult_2,]
Mult_2_R508B$Refrigerant <- "R508B"

Reefers_single <- Reefers_single_round1[-Mult_2,]
Reefers_single <- bind_rows(Reefers_single, Mult_2_R134A, Mult_2_R404A, Mult_2_R170, Mult_2_R508B)

# Confirm each entry only has one refrigerant and identify unique refrigerants
Refs_reefers <- unique(Reefers_single$Refrigerant)
Refs_misc <- unique(MiscApps_with$Refrigerant)
Refs_used <- unique(c(Refs_reefers, Refs_misc))

# Build a reference table in Excel so there's a record and it can be loaded in
```

```{r, echo = FALSE}
# Let's clean up to avoid mistakes
# Comment this out or edit it if you want to preserve more or different variables
rm(list=ls()[! ls() %in% c("Reefers_single", "MiscApps_with")])
```


```{r, echo = FALSE}
# Read in the refrigerant GWP info
RefrigerantGWP <- read_csv(here("Data", "RefrigerantGWP.csv"),
                           show_col_types = FALSE)
```

```{r, echo = FALSE}
# Calculate emissions for each category of refrigerator

# Domestic Refrigeration
Domestic_Reefers <- Reefers_single |> filter(Type != "Ultra Freezer" & (Note != "Commercial" | is.na(Note)))
# Add a column with the refrigerant GWP, to make summing across units easier
Domestic_Reefers_GWP <- full_join(Domestic_Reefers, RefrigerantGWP, by = "Refrigerant")
Domestic_Reefers_GWP <- Domestic_Reefers_GWP |> filter(!is.na(Type))
# Sort out mini fridges and mini freezers, NAs are a pain...
Mini_detect <- str_detect(Domestic_Reefers_GWP$Note, fixed('mini', ignore_case = TRUE))
Mini_idx <- which(Mini_detect == 'TRUE')
Reg_idx <- which(Mini_detect == 'FALSE' | is.na(Mini_detect))
Domestic_Reefers_GWP_minis <- Domestic_Reefers_GWP[Mini_idx,]
Domestic_Reefers_GWP_reg <- Domestic_Reefers_GWP[Reg_idx,]
DR_OppCap_mini <- unname(quantile(c(0.05,0.5)))[2]  # 'unname' removes the quantile names, the second is 25%
DR_OppCap_reg <- (0.05 + 0.5) / 2  # Operating capacity
DR_emissions_mini <- sum(DR_OppCap_mini * (0.5 / 100) * Domestic_Reefers_GWP_minis$GWP)/1000
DR_emissions_reg <- sum(DR_OppCap_reg * (0.5 / 100) * Domestic_Reefers_GWP_reg$GWP)/1000
DR_emissions <- DR_emissions_mini + DR_emissions_reg

# Stand-Alone Commercial
Commercial_Reefers <- Reefers_single |> filter(Type == "Ultra Freezer" | Note == "Commercial")
# Add a column with the refrigerant GWP, to make summing across units easier
Commercial_Reefers_GWP <- full_join(Commercial_Reefers, RefrigerantGWP, by = "Refrigerant")
Commercial_Reefers_GWP <- Commercial_Reefers_GWP |> filter(!is.na(Type))
CR_OppCap <- (0.2 + 6) / 2
CR_emissions <- sum(CR_OppCap * (15/100) * Commercial_Reefers_GWP$GWP)/1000

# Air conditioner (residential/commercial/AC)
AC <- MiscApps_with |> filter(Note == "Back-up Air conditioning unit")
# Add a column with the refrigerant GWP, to make summing across units easier
AC_GWP <- full_join(AC, RefrigerantGWP, by = "Refrigerant")
AC_GWP <- AC_GWP |> filter(!is.na(Type))
AC_OppCap <- 0.5
AC_emissions <- (AC_OppCap * (10/100) * AC_GWP$GWP)/1000

# Dehumidifiers (domestic refrigeration)
Dehumid <- MiscApps_with |> filter(Note == "Dehumidifier")
# Add a column with the refrigerant GWP, to make summing across units easier
Dehumid_GWP <- full_join(Dehumid, RefrigerantGWP, by = "Refrigerant")
Dehumid_GWP <- Dehumid_GWP |> filter(!is.na(Type))
DH_OppCap <- 0.05
DH_emissions <- sum(DH_OppCap * (0.5 / 100) * Dehumid_GWP$GWP)/1000

# Ice makers (stand-alone commercial)
IceMaker <- MiscApps_with |> filter(Note == "Ice Maker")
# Add a column with the refrigerant GWP, to make summing across units easier
IceMaker_GWP <- full_join(IceMaker, RefrigerantGWP, by = "Refrigerant")
IceMaker_GWP <- IceMaker_GWP |> filter(!is.na(Type))
IM_OppCap <- 0.2
IM_emissions <- (IM_OppCap * (15/100) * IceMaker_GWP$GWP)/1000

# Sum across categories
Reefer_Emissions <- DR_emissions + CR_emissions + AC_emissions + DH_emissions
```

