---
title: "PIFSC_NOAAships"
author: "Phoebe.Woodworth-Jefcoats@noaa.gov"
date: "2024-03-01"
output: html_document
---

# PIFSC use of NOAA ships - emissions inventory
This document describes the methodology used to calculate the carbon footprint of PIFSC's use of NOAA's ships.

### Data gathering
Data on fuel type and use, as well as miles traveled, were provided by the ships' operations officers.  Each ship provided slightly different data.  Full details are retained in copies of the emails accompanying the data (in our Google Drive).  Relevant details are included here.

Data provided by the NOAA Ship *Oscar Elton Sette*, per project:
- Total gallons of diesel fuel consumed
- Total nautical miles traveled
- Days in range
- Average fuel use per day
- Average mileage per day
- Day trip code, with *PDAS* identifying the periods of interest to this project
These data were provided in a spreadsheet named `SE fuel report.xlsx`.

Data provided by the NOAA Ship *Rainier*, per day:
- Total nautical miles traveled
- Total gallons of diesel fuel consumed
These data were provided in a spreadsheet named `NMR & Fuel Tracker 2023.xlsx`.
Relevant cruise dates were obtained from Jenni Samson (jennifer.samson@noaa.gov).  The *Rainier* data include fuel used for small boats, which averages about 30 gallons per day (see email from *Rainier* operations officer).

### Data assumptions
The following assumptions were made about our vehicles:  
- Miles were assumed to be nautical miles unless stated otherwise.  

### Data wrangling
##### *Sette*
Data for the *Sette* were loaded, filtered to PIFSC-only (`Day Type Code == PDAS`), and `Total Fuel (gals)` and `Total miles` were each summed.  These summed values were input into the EPA calculator tool.

##### *Rainier*
Data for the *Rainier* were loaded, filtered to the correct sea days, and fuel and miles were each summed.  These summed values were input into the EPA calculator tool

## Preliminary results
The preliminary estimate is that in FY23, PIFSC's use of NOAA ships emitted the equivalent of 2592.8 metric tons of CO~2~.

## Final estimate
The final estimate is that in FY23, PIFSC's use of NOAA ships emitted the equivalent of 2592.8 metric tons of CO~2~.

## Error checking
The data were also summed in a copy of the excel spreadsheets for comparison.  The values agreed with those calculated here.

The data supplied by the NOAA Ship *Reuben Lasker* included an estimate of pounds of CO~2~ emitted per gallon of fuel (pounds CO~2~ = gallons diesel × 22).  There are approximately 2205 pounds in a metric ton.  From our calculator:
248804 × 22 / 2205 = 2482.398 metric tons of CO~2~
This is in the ballpart of the spreadsheet calculator.

```{r}
library(tidyverse)
library(here)
library(readxl)

# SETTE
# Load data
# This will trigger a warning for the text summary at the end of numeric columns.
# You can ignore it.
Sette <- read_xlsx(here("Data", "SE fuel report.xlsx"), 
                        col_types = c("text", "text", "text", "text", "text", 
                                      "numeric", "numeric", "numeric", "numeric", "numeric"))

# Filter to only PIFSC sea days:
Sette_PIFSC <- filter(Sette, `Day Type Code` == 'PDAS')

# Sum fuel and mileage
Sette_PIFSC_fuel <- summarise(Sette_PIFSC, `Total Fuel (gals)` = sum(`Total Fuel (gals)`))
Sette_PIFSC_miles <- summarise(Sette_PIFSC, `Total miles` = sum(`Total miles`))

# Convert nautical miles to statute miles
Sette_PIFSC_miles = Sette_PIFSC_miles * 1.15

# RAINIER
# Load data
# The column with the name of the month reflects the day of the month
# The column to the immediate right of the one with the name of the month reflects the value:
# either miles run or fuel consumed
# Also, the fuel data contain a mix of 0 and NA, with no fuel use in Nov or Dec
Rainier_miles_sheet <- read_xlsx(here("Data", "NMR & Fuel Tracker 2023.xlsx"), 
                                 sheet = "NMR", skip = 5)
Rainier_fuel_sheet <- read_xlsx(here("Data", "NMR & Fuel Tracker 2023.xlsx"), 
                                sheet = "FUEL", skip = 5)

# Each sheet has an empty row under the column name - remove it
# Now the row numbers match the days of the month
Rainier_miles_sheet <- Rainier_miles_sheet[rowSums(is.na(Rainier_miles_sheet)) != 
                                             ncol(Rainier_miles_sheet),]
Rainier_fuel_sheet <- Rainier_fuel_sheet[rowSums(is.na(Rainier_fuel_sheet)) != 
                                           ncol(Rainier_fuel_sheet),]

# From Jenni's email, identify the correct days
Rainier_miles <- sum(Rainier_miles_sheet$...6[9:23],   # Leg 2, March
                     Rainier_miles_sheet$...6[29:31],  # Leg 3, March
                     Rainier_miles_sheet$...8[1:21],   # Leg 3, April
                     Rainier_miles_sheet$...8[25:30],  # Leg 4, April
                     Rainier_miles_sheet$...10[1:13],  # Leg 4, May
                     Rainier_miles_sheet$...12[19:30], # Leg 5/6, June
                     Rainier_miles_sheet$...14[1:12],  # Leg 5/6, July
                     Rainier_miles_sheet$...14[21:31], # Leg 7, July
                     Rainier_miles_sheet$...16[1:10],  # Leg 7, August
                     Rainier_miles_sheet$...18[4:5],   # Transit
                     na.rm = TRUE) 

Rainier_fuel <- sum(Rainier_fuel_sheet$...6[9:23],     # Leg 2, March 
                    Rainier_fuel_sheet$...6[29:31],    # Leg 3, March
                    Rainier_fuel_sheet$...8[1:21],     # Leg 3, April
                    Rainier_fuel_sheet$...8[25:30],    # Leg 4, April
                    Rainier_fuel_sheet$...10[1:13],    # Leg 4, May
                    Rainier_fuel_sheet$...12[19:30],   # Leg 5/6, June
                    Rainier_fuel_sheet$...14[1:12],    # Leg 5/6, July
                    Rainier_fuel_sheet$...14[21:31],   # Leg 7, July
                    Rainier_fuel_sheet$...16[1:10],    # Leg 7, August
                    Rainier_fuel_sheet$...18[4:5],     # Transit
                    na.rm = TRUE) 

# Convert from nautical to statute miles
Rainier_miles <- Rainier_miles * 1.15
```

```{r}
# Let's estimate things like gallons per mile
# From Jenni's email, identify the correct days
Rainier_daily_miles <- rbind(as.matrix(Rainier_miles_sheet$...6[9:23]),   # Leg 2, March
                     as.matrix(Rainier_miles_sheet$...6[29:31]),  # Leg 3, March
                     as.matrix(Rainier_miles_sheet$...8[1:21]),   # Leg 3, April
                     as.matrix(Rainier_miles_sheet$...8[25:30]),  # Leg 4, April
                     as.matrix(Rainier_miles_sheet$...10[1:13]),  # Leg 4, May
                     as.matrix(Rainier_miles_sheet$...12[19:30]), # Leg 5/6, June
                     as.matrix(Rainier_miles_sheet$...14[1:12]),  # Leg 5/6, July
                     as.matrix(Rainier_miles_sheet$...14[21:31]), # Leg 7, July
                     as.matrix(Rainier_miles_sheet$...16[1:10]),  # Leg 7, August
                     as.matrix(Rainier_miles_sheet$...18[4:5]))   # Transit

Rainier_daily_fuel <- rbind(as.matrix(Rainier_fuel_sheet$...6[9:23]),     # Leg 2, March 
                    as.matrix(Rainier_fuel_sheet$...6[29:31]),    # Leg 3, March
                    as.matrix(Rainier_fuel_sheet$...8[1:21]),     # Leg 3, April
                    as.matrix(Rainier_fuel_sheet$...8[25:30]),    # Leg 4, April
                    as.matrix(Rainier_fuel_sheet$...10[1:13]),    # Leg 4, May
                    as.matrix(Rainier_fuel_sheet$...12[19:30]),   # Leg 5/6, June
                    as.matrix(Rainier_fuel_sheet$...14[1:12]),    # Leg 5/6, July
                    as.matrix(Rainier_fuel_sheet$...14[21:31]),   # Leg 7, July
                    as.matrix(Rainier_fuel_sheet$...16[1:10]),    # Leg 7, August
                    as.matrix(Rainier_fuel_sheet$...18[4:5]))     # Transit

GallonsPerMile <- Rainier_daily_fuel/Rainier_daily_miles
GallonsPerMile <- GallonsPerMile[which(is.finite(GallonsPerMile) == TRUE)]  #Dividing by zero is mathematically problematic

AvgGalPerMile <- mean(GallonsPerMile)

# Gallons per mile * 20 nm roundtrip * 22 lbs per gallon / 2205 lbs in a metric ton
EmissionsSaved <- AvgGalPerMile * 20 * 22 / 2205
```

```{r}
# Looking at data from the Lasker.  Even though it's not from FY23, their Ops had some handy insights 
# It would be cool if we could quantify them.

# Load data
HICEAS1 <- read_xlsx(here("Data", "HICEAS Reuben Lasker Fuel Efficiency Tracker.xlsx"),
                   sheet = "HICEAS 2023 Leg 1", skip = 1)

HICEAS2 <- read_xlsx(here("Data", "HICEAS Reuben Lasker Fuel Efficiency Tracker.xlsx"),
                   sheet = "HICEAS 2023 Leg 2", skip = 1)

Lasker_Miles <- rbind(as.matrix(HICEAS1$Miles), 
                      as.matrix(HICEAS2$Miles))
Lasker_Gallons <- rbind(as.matrix(HICEAS1$Gallons), 
                        as.matrix(HICEAS2$Gallons))

Lasker_GallonsPerMile <- Lasker_Gallons/Lasker_Miles

```

```{r}
# Count days at sea to estimate max savings of dropping one generator
Sette_Days <- sum(Sette_PIFSC$`Days in Range`)
Rainier_Days <- as.numeric(dim(Rainier_daily_miles)[1])

# Very rough estimate from the Sette's 'Ship Energy Efficiency Management Plan'
# is that going down by a single generator saves maybe 5 gal per hour
# 5 gallons per hour * 24 hours in a day * # of days
FuelSvgMax <- (Sette_Days + Rainier_Days) * 5 * 24

# 22 lbs per gallon / 2205 lbs in a metric ton
EmissingsSvgMax <- FuelSvgMax * 22 / 2205
```